---
title: "Dimensionality reduction and clustering"
author: "*Lisa Loksø Dietz*"
date: "*`r format(Sys.time(), '%Y-%m-%d')`*"
knit: (function(inputFile, encoding) { 
      out_dir = paste0(dirname(getwd()),"/3_output");
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file=file.path(out_dir, '5_dimred_clustering.html')) })
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    toc_collapsed: true
    code_folding: show
---

```{r Setup 1, include = FALSE}

knitr::opts_chunk$set(fig.width=10, fig.height=7, out.width = "70%",fig.align = 'center', eval=TRUE) 
# eval controls whether code should be run when knitting
# cache will avoid running again the code chunk again if you didn't change anything within it

```

## 0) Load source and set dir

```{r Directories, message = FALSE, warning = FALSE}

rm(list=ls())
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
source("0_source.R")

```

## 1) Load data

### 1.1) Cleaned data

This is the experiment metadata.

```{r Read files, message = FALSE}

read_dir = glue("{dir_project}/2_pipeline/{folder_QC}/out")
fs_clean = readRDS(glue("{read_dir}/FlowSet_clean.rds"))
pData(fs_clean)

```

### 1.2) Spectral flow panel

This is the panel used. Type markers are markers that indicate surface markers, such as CD3, CD4, or markers that you do want to use for clustering. State markers are markers that indicate "state" of a cell, such as PD-L1 marker, or use it to indicate markers you won't use for clustering.

```{r Read panel, message = FALSE}

# read panel
panel = read.csv(file = glue("{dir_project}/2_pipeline/{folder_QC}/out/Panel.csv"))[,-1]

# define type and state markers
gated_markers = c("CD14_16","CD19","CD3")
features_state= panel[panel$marker_class=="state","antigen"]
features_type = setdiff(panel[panel$marker_class=="type","antigen"], gated_markers)
features_effectormemory = c("CD45RA","CCR7","CD27","CD95")
features_inhibitory = c("TIGIT","PD1","CD39")
features_chemokineR = c("CCR4","CXCR5")
features_cytokine_degran = c("CD107a","IFN","TNF","IL2","IL413")
features_effectorproteins = c("Perforin","GzmB","GzmK","Granulysin")
features_TF = c("Tbet","TCF1","FoxP3")
features_prolif = c("Ki67")
features_clustering = c(features_type, features_effectormemory, "FoxP3")

# choose channels to use in SCE
panel[panel$fcs_colname == "SSC-A", "antigen"] = "SSC"
panel[panel$fcs_colname == "SSC-A", "marker_class"] = "type"
channels_sce = panel[(panel$marker_class!="none")&(!panel$antigen %in% gated_markers),"fcs_colname"]

total_markers = length(channels_sce)

```

## 2) Clustering and visualization

### 2.1) Single Cell Experiment (SCE)

```{r Create SCE, message = FALSE}

set.seed(1234)
sce = prepData(fs_clean,
               md = pData(fs_clean),
               panel = panel,
               features = channels_sce,
               FACS = TRUE,
               transform = FALSE,
               panel_cols = list(channel = "fcs_colname",
                                 antigen = "antigen"
                                 ),
               md_cols = list(file = "name",
                              id = "name",
                              factors = c("stimulation"
                                          )
                              )
               )

assayNames(sce)[1] = "exprs"

saveRDS(sce, file = glue("{dir_store}/sce.rds"))
sce = readRDS(file = glue("{dir_store}/sce.rds"))

sce


```

### 2.2) Cluster SCE and delta area plot

```{r Cluster and delta area plot, message = FALSE}

set.seed(1234)
no_clusters = 25
meta_name = glue("meta{no_clusters}")
features_clustering

sce = cluster(sce,
              features = features_clustering,
              maxK = no_clusters,
              xdim = 10,
              ydim = 10,
              verbose = TRUE,
              seed = 1234)

delta_area(sce)

```

```{r Save clustered data, message = FALSE}

saveRDS(sce, file = glue("{dir_store}/sce_clust.rds"))
sce = readRDS(file = glue("{dir_store}/sce_clust.rds"))
sce

```


### 2.3) Visualization

**Minimum spanning tree (MST)**


```{r Plot MST, message = FALSE}
no_clusters = 25
channels_fsom = panel[panel$antigen %in% features_clustering,"fcs_colname"]

# specify grid & markers for clustering
fsom = FlowSOM::FlowSOM(input = fs_clean,
                        scale = F,
                        colsToUse = channels_fsom, 
                        seed = 1234,
                        nClus = no_clusters, 
                        xdim = 10,
                        ydim = 10)

mst = FlowSOM::BuildMST(fsom)

# visualize with 'PlotStars'
png(file = glue("{dir_store}/fsom_meta{no_clusters}.png"), width=14,height=12,units="in",res=400)
FlowSOM::PlotStars(fsom,
          backgroundValues = fsom$metaclustering,
          backgroundColors = colorRampPalette(CATALYST:::.cluster_cols)(25))
dev.off()
#FlowSOM::PlotNumbers(fsom)

```

**Number of cells per sample**

The plots show cell count for each of the samples on top and the relative abundance of each matched setnumber:

```{r plotCounts, message = FALSE}

plotCounts(sce,
           group_by = "sample_id",
           color_by = "stimulation")

```

**Pseudobulk-level MDS plot**

A pbMDS plot will give a sense of similarities between clusters and/or samples in an unsupervised way and of key difference in expression before conducting any formal testing.

**MDS on sample-level pseudobulks**

```{r Sample-level pseudobulks, message = FALSE, warning=FALSE, fig.width=5, fig.height=5}

ei(sce)$stimulation

# sample-level pseudobulks
pbMDS(sce,
      features = "type",
      k = "meta20",
      by = "both",
      shape_by = "stimulation",
      size_by = F)+ 
  theme(aspect.ratio = .7)

```

**MDS on pseudobulks by cluster-sample**

Pseudobulks computed for each cluster-sample colored by cluster (to highlight similarity between cell subpopulations), and shaped by condition (to reveal subpopulation-specific expression changes across conditions).

Here, see that cluster-sample instances of the same cell subpopulations group together. Meanwhile, most subpopulations exhibit a shift between instances where samples come from different infection groups.

```{r By cluster-sample, message = FALSE, warning=FALSE, fig.width=5, fig.height=5}
meta_name = "meta20"
# pseudobulks by cluster-sample
pbMDS(sce,
       features = "type",
       by = "both",
       k = "meta20",
       shape_by = "stimulation",
       size_by = FALSE)+ 
  theme(aspect.ratio = .7)


```

**Reduced dimension plot on CLR of proportions**

A dimensionality reduction plot on centered log-ratios (CLR) of sample/cluster proportions across clusters/samples will give a sense of similarities between samples/clusters in terms of their composition.

*CLR on cluster proportions across samples*

We here visualize the first two PCs computed on CLRs of sample proportions across clusters: small distances between samples mean similar cluster compositions between them, while large distances are indicative of differences in cluster proportions.

```{r Cluster proportions across samples, message = FALSE}

clrDR(sce,
      dr = "PCA",
      by = "sample_id",
      point_col = "stimulation",
      k = "meta20")

```

*CLR on sample proportions across clusters*

```{r Sample proportions across clusters, message = FALSE}

clrDR(sce,
      by = "cluster_id",
      arrow_col = "stimulation",
      size_by = FALSE) 

```

**Heatmap of aggregated marker expressions**

Heatmaps of aggregated (pseudobulk) marker expressions.

*Clustering markers*

```{r Type markers, message = FALSE}

plotExprHeatmap(sce,
                features = features_clustering,
                by = "sample_id",
                row_anno = c("stimulation"),
                scale = "last",
                bars = FALSE)

```

*Cytokines and degranulation+effector*

```{r Differentiation markers, message = FALSE}

plotExprHeatmap(sce,
                features = c(features_cytokine_degran,features_effectorproteins),
                by = "sample_id",
                row_anno = c("stimulation"),
                scale = "last",
                bars = FALSE)

```

*State markers*

```{r Activation markers, message = FALSE}

plotExprHeatmap(sce,
                features = features_state,
                by = "sample_id",
                row_anno = c("stimulation"),
                scale = "last",
                bars = FALSE)

```


**Pseudobulk expression boxplot**

*By metacluster*

By metacluster to compare medians for each cluster, and potentially identify changes across conditions early on:

```{r By metacluster, message = FALSE, fig.width=15, fig.height=12}

plotPbExprs(sce,
            k = "meta20",
            features = features_clustering,
            facet_by = "cluster_id",
            color_by = "stimulation")

```

*By antigen*

By antigen in order to compare marker expressions calculated over all cells across conditions:

```{r By antigen, message = FALSE, fig.width=15, fig.height=12}

plotPbExprs(sce,
            k = "meta20",
            features = features_clustering,
            facet_by = "antigen",
            color_by = "stimulation")

```

*Type-marker expression across clusters*

Aggregate expression values by both clusters and samples to see how consistent type-markers are expressed across clusters. The resulting plot gives an indication how good our selection of type markers is: Ideally, their expression should be fairly specific to a subset of clusters.

```{r Type-marker expression across clusters, message = FALSE, fig.width=15, fig.height=8}

plotPbExprs(sce,
            k = "meta20",
            features = features_clustering,
            group_by = "cluster_id",
            color_by = "sample_id",
            size_by = TRUE,
            geom = "points",
            jitter = FALSE)

```



### 2.4) Dimensionality reduction

Dimensionality reduction of the FlowSOM clusters in our SCE using the following markers:

```{r Run dimensionality reduction, message = FALSE}

set.seed(1234)
sce = runDR(sce,
            "UMAP",
            cells = NULL,
            features = features_clustering)

```


```{r Plot, message=FALSE}

plotDR(sce,
       color_by = "meta25")

```



### 2.5) Save dimred SCE 

```{r Save SCE, message = FALSE}

saveRDS(sce, glue("{dir_out}/sce_clust_dimred.rds"))

```

### 2.6) Read dimred SCE

```{r Read SCE, message = FALSE}

sce = readRDS(file = glue("{dir_out}/sce_clust_dimred.rds"))
sce$stimulation = factor(sce$stimulation,levels = c("NEG", "GAG", "SEB"))

```

## 3) Cluster visualization


### 3.1) Clustree and UMAPs

```{r Clustree, message = FALSE, fig.width = 20, fig.height = 14}

meta_range = c(2:25)
sce_plot = sce

png(file = glue("{dir_store}/clustree_meta{max(meta_range)}.png"), width=14,height=12,units="in",res=200)
q_Clustree(sce_plot, metarange = meta_range)
dev.off()

# list3 = list()
# for(metalevel in meta_range){
#   
#   meta_name = glue("meta{metalevel}")
#   
#   plt = q_plotDR(q_DownsampleSCE(sce_plot,1000),
#                  color_by = meta_name,
#                  labelsize = 3)+
#   ggtitle(label = glue("{meta_name}"))+
#   theme(plot.title = element_text(hjust = 0.5))
#   
#   list3[[meta_name]] = plt
#   
# }
# 
# png(file = glue("{dir_store}/UMAPs.png"), width=40,height=20,units="in",res=200)
# grid.arrange(grobs = list3, ncol = 5)
# dev.off()


```

### 3.2) Visualizing FlowSOM clusters on UMAP projection

```{r Visualize FlowSOM clusters on UMAP, message = FALSE, fig.width=14, fig.height=12}

width = 14
height = 12

width = 14*0.5
height = 12*0.5

png(file = glue("{dir_store}/UMAP_meta25.png"), width=width,height=height,units="in",res=200)
q_plotDR(sce,
         color_by = "meta25",
         labelsize = 5)
dev.off()

png(file = glue("{dir_store}/UMAP_facet_stim_meta25.png"), width=24,height=height,units="in",res=200)
q_plotDR(sce,
         color_by = "meta25",
         labelsize = 5,
         facet_by = "stimulation")
dev.off()

png(file = glue("{dir_store}/UMAP_clust.png"), width=12,height=5,units="in",res=200)
plotDR(sce,
       color_by = features_clustering,
       ncol = 4)
dev.off()

width = 5
height = 4.29

png(file = glue("{dir_store}/UMAP_meta25_nolabel.png"), width=width,height=height,units="in",res=200)
plotDR(sce,
         color_by = "meta25")
dev.off()


for(feature in features_clustering){
  png(file = glue("{dir_store}/UMAP_{feature}.png"), width=width,height=height,units="in",res=200)
  print(plotDR(sce,
         color_by = feature))
  dev.off()
}

plot_list = list()
for(feature in c(features_type,features_state)){
  p = plotDR(sce,
         color_by = feature,
         facet_by = "stimulation")
  plot_list[[feature]] = p
}

png(file = glue("{dir_store}/UMAP_facet_stim_all.png"), width=14,height=100,units="in",res=200)
grid.arrange(grobs = plot_list, ncol = 1)
dev.off()

for(feature in c(features_type,features_state)){
  png(file = glue("{dir_store}/UMAP_facet_stim_{feature}.png"), width=14,height=12,units="in",res=200)
  print(plotDR(sce,
         color_by = feature,
         facet_by = "stimulation"))
  dev.off()
}



png(file = glue("{dir_store}/density_meta25.png"), width = 16, height = 14, units = "in", res=200)
plotClusterExprs(sce,
                 k = "meta25",
                 features = features_clustering)
dev.off()


```


### 3.3) Add manual labels

```{r Load SCE, message = FALSE}

aggregate_labels = readRDS(glue("{dir_project}/2_pipeline/{folder_manualgate}/out/aggregate_labels.rds"))
manual_labels_SCE = as.factor(aggregate_labels)
sce$manual_id = manual_labels_SCE

```


### 3.4) Visualizing manual labels on UMAP

```{r Add manual cell labels, message = F, fig.width = 20, fig.height = 20}

cell_types_of_interest <- setdiff(levels(sce$manual_id),"Unlabeled")
width = 14*0.8
height = 12*0.8
png(filename = glue("{dir_store}/UMAP_manual.png"),width = width, height = height, units = "in", res = 200)
q_plotDR(sce,
         color_by = "manual_id",
         labelsize = 5,
         highlight_clust = cell_types_of_interest)
dev.off()

png(filename = glue("{dir_store}/UMAP_manual_facet_stim.png"),width = 24, height = 12, units = "in", res = 200)
q_plotDR(sce,
         color_by = "manual_id",
         labelsize = 5,
         highlight_clust = cell_types_of_interest,
         facet_by = "stimulation")
dev.off()

cell_types_of_interest <- setdiff(levels(sce$manual_id),"Unlabeled")
png(filename = glue("{dir_store}/UMAP_manual.png"),width = 14, height = 12, units = "in", res = 200)
q_plotDR(sce,
         color_by = "manual_id",
         labelsize = 5,
         highlight_clust = cell_types_of_interest)
dev.off()

```


### 3.4) Add boolean matrix of marker positivity

```{r}

boolean_matrix_marker_positivity = read.csv(file = glue("{dir_project}/2_pipeline/6_gateMarkers/out/boolean_matrix_markerpositivity.csv"))[,-1]
names(boolean_matrix_marker_positivity) = gsub(".$","_pos",names(boolean_matrix_marker_positivity))
marker_positivity_to_plot = names(boolean_matrix_marker_positivity)[-1]
new_coldata = cbind(colData(sce),boolean_matrix_marker_positivity)
sce = scdrake::sce_add_colData(sce, new_coldata, replace = TRUE)

image(as.matrix(boolean_matrix_marker_positivity))
colnames(boolean_matrix_marker_positivity)

```

### 3.5) Plot UMAP of boolean markers

```{r}

plot_list = list()
for(color_plot in marker_positivity_to_plot){
  p = plotDR(sce,
       color_by = color_plot,
       facet_by = "stimulation")+
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "#00FF00"))
  plot_list[[color_plot]] = p
}

png(file = glue("{dir_store}/UMAP_facet_stim_all_pos.png"), width=14,height=100,units="in",res=200)
grid.arrange(grobs = plot_list, ncol = 1)
dev.off()


for(color_plot in marker_positivity_to_plot){
  
  png(filename = glue("{dir_store}/UMAP_facet_stim_{color_plot}.png"), width=14,height=12,units="in",res=200)
  print(plotDR(sce,
       color_by = color_plot,
       facet_by = "stimulation")+
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "#00FF00")))
  dev.off()
  
}

q_plotDR_shuffled(sce,
                  color_by = "IFN_pos",
                  facet_by = "stimulation")

```

### 3.6) Backgroup subtracted (expression)

```{r Make dataframe of sce, message = F}

colData(sce)
names(cluster_codes(sce))
assayNames(sce)
#assay(sce,"exprs")
rownames(assay(sce,"exprs"))
colData(sce)$meta25 = cluster_ids(sce,"meta25")

rowsum(colData(sce),group = c("stimulation","meta25"))

df = t(assay(sce,"exprs"))
df = cbind(df,colData(sce))
df = data.frame(df)
head(df)
colnames(df)
df$stimulation = factor(df$stimulation,levels = c("NEG","GAG","SEB"))

write.csv(df, file = glue("{dir_out}/dataframe_sce.csv"))

df <- df %>% select(-contains('_pos'))

df_by_meta25_stim = aggregate(. ~ meta25+stimulation, data=df[,!(names(df) %in% c("cluster_id", "manual_id","sample_id"))], FUN=median)

names(df_by_meta25_stim)

df_long = df_by_meta25_stim %>% 
  pivot_longer(cols = -c(meta25, stimulation), 
               names_to = "marker",
               values_to = "value")

df_long = df_long %>% 
   group_by(meta25) %>% 
   mutate(value_backgroundsubtracted = value- value[stimulation == "NEG"])

df_long[(df_long$meta25 == 1 & df_long$marker=="IL2"),]

pdf(file = glue("{dir_store}/markers_backgroundsubtracted.pdf"), width = 10,height=90)
ggplot(df_long, aes(fill=stimulation, y=value_backgroundsubtracted, x=meta25)) + 
  geom_bar(position="dodge", stat="identity")+
  facet_wrap(~marker,ncol = 1, scales="free_y")+
  scale_fill_manual(values = c("NEG" = "grey70", "GAG" = "#0bb4ff","SEB"="#e60049"))
dev.off()

features1 = features_clustering
features2 = c(features_cytokine_degran,features_chemokineR)
features3 = features_effectorproteins
features4 = features_inhibitory


```


### 2.7) Background subtracted (percent pos)

```{r}

n_cells_sce = nrow(colData(sce))
n_markers_gated = length(which(grepl(pattern = "*_pos",names(colData(sce)))))-1

# add meta25 to sce
colData(sce)$meta25 = cluster_ids(sce, "meta25")

# make dataframe from sce coldata (rows = n cells in sce)
df = data.frame(colData(sce)[,!(names(colData(sce)) %in% c("cluster_id", "manual_id","sample_id","roo_pos"))])

#check lenght of df
nrow(df) == n_cells_sce

# make the df long format (rows = n cells in sce * 18 gated markers = 2,472,912)
df_long = df %>% 
  pivot_longer(cols = -c(meta25, stimulation), 
               names_to = "marker",
               values_to = "pos")
# check lenght
nrow(df_long) == nrow(colData(sce))*18

# group by meta25, stimulation, and marker and get number of positive cells
counts_pos = df_long %>% 
  group_by(meta25, stimulation,marker) %>% 
  summarise(
    n_pos = sum(pos))

# group by meta25 and stimulation to get number of cells per cluster per stimulation
counts_cluster = df %>% 
  group_by(meta25, stimulation) %>% 
  tally()

# check the sum is equal to n cells in sce
sum(counts_cluster$n) == n_cells_sce

# check visually that the cluster sizes per stimulation are correct
# ggplot(counts_cluster,
#        aes(x=meta25,y=n,group = stimulation,fill=stimulation))+
#   geom_bar(position="dodge", stat="identity")
# 
# plotCounts(sce, group_by = "meta25", color_by = "stimulation", prop = FALSE)+
#   scale_fill_manual(values = c("NEG" = "grey70", "GAG" = "#0bb4ff","SEB"="#e60049"))+
#   scale_x_continuous(breaks = c(1:25))

# combine the counts of positive cells with the cluster sizes (nrow = 25 * 3* 18)
df_counts = counts_pos %>% right_join(counts_cluster, by=c("meta25","stimulation"))
# check lenght 
nrow(df_counts) == 25*3*18
# calculate pers_pos
df_counts$perc_pos = (df_counts$n_pos/df_counts$n)*100

# subtract negative value for same cluster and same marker
df_final = df_counts %>% 
   group_by(meta25) %>% 
   mutate(perc_pos_backgroundsubtracted = perc_pos- perc_pos[stimulation == "NEG"])

# check all negative are 0
sum(df_final[df_final$stimulation=="NEG","perc_pos_backgroundsubtracted"])

# remove all stimulation == NEG
df_final = df_final[df_final$stimulation!="NEG",]

# pdf(file = glue("{dir_store}/percpos_backgroundsubtracted.pdf"), width = 10,height=90)
# ggplot(df_final, aes(fill=stimulation, y=percpos_backgroundsubtracted, x=meta25)) + 
#   geom_bar(position="dodge", stat="identity")+
#   facet_wrap(~marker,ncol = 1, scales="free_y")+
#   scale_fill_manual(values = c("NEG" = "grey70", "GAG" = "#0bb4ff","SEB"="#e60049"))+scale_y_continuous(trans=scales::pseudo_log_trans(base = 10))
# dev.off()

# recalculate number of positive cells by multiplying per_pos with n cells (--> gets a result relative to cluster size)
df_final$n_pos_backgroundsubtracted = (df_final$perc_pos_backgroundsubtracted/100)*df_final$n

# set negative values to zero (if not, the pie chart looks weird with blank space)
df_final$n_pos_backgroundsubtracted = pmax(df_final$n_pos_backgroundsubtracted,0)

# df_final[(df_final$meta25==16 & df_final$marker %in% c("CD107a_pos","IFN_pos","TNF_pos")),c("meta25","stimulation","marker","n","n_pos","perc_pos","perc_pos_backgroundsubtracted","n_pos_backgroundsubtracted")]

# ggplot(df_final[(df_final$marker %in% c("TNF_pos","IFN_pos","CD107a_pos")&df_final$stimulation=="GAG"),], 
#        aes(fill=stimulation,
#            y=n_pos_backgroundsubtracted, 
#            x=meta25)) + 
#   geom_bar(position="dodge", 
#            stat="identity",
#            show.legend=F)+
#   facet_wrap(~marker,
#              ncol = 3, 
#              scales="free_y",
#              labeller = as_labeller(c("CD107a_pos" = "CD107a",
#                                     "IFN_pos" = "IFN",
#                                     "TNF_pos" = "TNF")))+
#   scale_fill_manual(values = c("GAG" = "red"))+
#   ylab("Percentage positive cells in cluster (NEG stimulation subtracted)")+
#   theme_minimal()
# 
# df_plot = df_final[(df_final$marker %in% c("TNF_pos")&df_final$stimulation=="GAG"),c("meta25","n_pos_backgroundsubtracted")]
# df_plot$n_pos_backgroundsubtracted = as.numeric(df_plot$n_pos_backgroundsubtracted)
# 
# ggplot(df_plot, aes(x="", y=n_pos_backgroundsubtracted, fill=meta25))+
#   geom_bar(width = 1, stat = "identity")+
#   coord_polar("y", start=0)+
#   theme_minimal()+
#   scale_fill_manual(values = CATALYST:::.cluster_cols[1:25])

# define markers to plot
markers_to_plot = c("TNF_pos","IFN_pos","CD107a_pos")
df_plot = df_final[(df_final$marker %in% markers_to_plot &df_final$stimulation=="GAG"),c("meta25","marker","n_pos_backgroundsubtracted")]

# transform n_pos_backgroundsubtracted to a percentage 
df_plot = df_plot %>% group_by(marker) %>%
  mutate(percentage = n_pos_backgroundsubtracted / sum(n_pos_backgroundsubtracted) * 100)

# # get positions for labels
# df2 <- df_plot %>% 
#   mutate(csum = rev(cumsum(rev(percentage))), 
#          pos = percentage/2 + lead(csum, 1),
#          pos = if_else(is.na(pos), percentage/2, pos))

# TODO: below 10 % --> group together in other (describe what's in other)
png(filename = glue("{dir_store}/pie_charts.png"),height = 10,width = 20,units = "in",res = 300)
ggplot(df_plot, aes(x="", y=percentage, fill=meta25))+
  geom_bar(width = 1, stat = "identity")+
  facet_wrap(~marker)+
  # geom_label_repel(data = df2,
  #                  aes(y = pos, label = glue("meta{meta25} ({round(percentage,2)} %)")),
  #                  size = 3, nudge_x = 1, show.legend = FALSE) +
  scale_fill_manual(values = CATALYST:::.cluster_cols[1:25])+coord_polar("y",start=0)+
  blank_theme+theme(axis.text.x=element_blank())
dev.off()


blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank()
  )

```


## 4) Plot expression

```{r Marker expression between conditions, message = F}

palette = values = c("NEG" = "grey70", "GAG" = "#0bb4ff","SEB"="#e60049")
features1 = features_clustering
features2 = c(features_cytokine_degran,features_chemokineR)
features3 = features_effectorproteins
features4 = features_inhibitory

png(filename = glue("{dir_store}/markerexpr_clustering.png"),width = 12, height = 6,unit = "in",res = 300)
plotPbExprs(sce,
            k = "meta25",
            features = features1,
            facet_by = "antigen",
            color_by = "stimulation",
            group_by = "cluster_id",
            geom = "point",
            jitter=F)+
  scale_color_manual(values = palette)+
  ggtitle(label = "Markers used for clustering")
dev.off()

png(filename = glue("{dir_store}/markerexpr_cytokines_chemokines.png"),width = 12, height = 10,unit = "in",res = 300)
plotPbExprs(sce,
            k = "meta25",
            features = features2,
            facet_by = "antigen",
            color_by = "stimulation",
            group_by = "cluster_id",
            geom = "point",
            jitter=F)+
  scale_color_manual(values = palette)+
  ggtitle(label = "Cytokines and chemokines")
dev.off()

png(filename = glue("{dir_store}/markerexpr_effectormolecules.png"),width = 8, height = 6,unit = "in",res = 300)
plotPbExprs(sce,
            k = "meta25",
            features = features3,
            facet_by = "antigen",
            color_by = "stimulation",
            group_by = "cluster_id",
            geom = "point",
            jitter=F)+
  scale_color_manual(values = palette)+
  ggtitle(label = "Effector molecules")
dev.off()

png(filename = glue("{dir_store}/markerexpr_inhibmolecules.png"),width = 12, height = 4,unit = "in",res = 300)
plotPbExprs(sce,
            k = "meta25",
            features = features4,
            facet_by = "antigen",
            color_by = "stimulation",
            group_by = "cluster_id",
            geom = "point",
            jitter=F)+
  scale_color_manual(values = palette)+
  ggtitle(label = "Inhibitory molecules")
dev.off()


```

## 5) Save flowSet with cluster ids

```{r save flowSet, message = F}

colData(sce)
colData(sce)$meta25 = cluster_ids(sce,"meta25")
colData(sce)
fs_clust = sce2fcs(sce, 
                   split_by = "sample_id", 
                   keep_cd = T, 
                   keep_dr = F, 
                   assay = "exprs")

#flowCore::fr_append_cols(ff_o, m)

saveRDS(fs_clust, glue("{dir_out}/flowSet_clust.rds"))

```

```{r Extract clusters, message=FALSE}

sce_NEG = filterSCE(sce,
                    stimulation == "NEG")

sce_GAG = filterSCE(sce,
                    stimulation == "GAG")

sce_SEB = filterSCE(sce,
                    stimulation == "SEB")

extractClusters(sce_NEG,
                                 k = "meta25",
                                 clusters = as.character(1),
                                 as = "flowSet")

ei(sce)
# extractClusters does not conserve the file ID
my_list = list()
list_clusters = list()
for(stim in c(ei(sce)$stimulation)){
  
  print(stim)
  sce_filtered = filterSCE(sce,
                           stimulation == stim)
  
  for(cluster_no in c(1:25)){
    
    name_metacluster = glue("meta{cluster_no}")
    name_flowSet = glue("flowSet_{name_metacluster}")
    
    fs_cluster = extractClusters(sce_filtered,
                                 k = "meta25",
                                 clusters = as.character(cluster_no),
                                 as = "flowSet",
                                 verbose = F)
    
    list_clusters[[name_metacluster]] = fs_cluster
  
  }
  
  my_list[[stim]] = list_clusters

}


saveRDS(my_list, file = glue("{dir_out}/list_flowSets.rds"))

```

# PROJECT DESCRIPTION FIG

```{r Figure for PhD, message = F}
cell_types_of_interest <- setdiff(levels(sce$manual_id),"Unlabeled")
blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank()
  )


fig1a = q_plotDR(sce,
                 color_by = "manual_id",
                 labelsize = 3,
                 highlight_clust = cell_types_of_interest)+ 
  guides(color=guide_legend(title="Manual gating",override.aes = list(size = 4)))


fig1b = q_plotDR(sce,
                 color_by = "meta25",
                 labelsize = 3)+ 
  guides(color=guide_legend(title="FlowSOM metacluster",override.aes = list(size = 4)))



fig1c = plotDR(sce,
               color_by = features_clustering,
               ncol = 2)

fig1d = plotDR(filterSCE(sce,
                         stimulation != "SEB"),
               color_by = "IFN_pos",
               facet_by = "stimulation",
               ncol = 2)+
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "red"))+
  guides(color=guide_legend(title="IFN positive",override.aes = list(size = 4)))

fig1d_alt = q_plotDR_shuffled(filterSCE(sce,
                         stimulation != "SEB"),
               color_by = "IFN_pos",
               facet_by = "stimulation",
               ncol = 1)+
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "red"))+
  guides(color=guide_legend(title="IFN positive",override.aes = list(size = 4)))

fig1d_alt$layers[[1]]$aes_params$size = 0.005

markers_to_plot = c("IFN_pos")
df_plot = df_final[(df_final$marker %in% markers_to_plot &df_final$stimulation=="GAG"),c("meta25","marker","n_pos_backgroundsubtracted")]

# transform n_pos_backgroundsubtracted to a percentage 
df_plot = df_plot %>% group_by(marker) %>%
  mutate(percentage = n_pos_backgroundsubtracted / sum(n_pos_backgroundsubtracted) * 100)

df_plot$marker = factor(df_plot$marker,levels = c("IFN_pos","TNF_pos","CD107a_pos"))

fig1e = print(ggplot(df_plot, 
               aes(x="", 
                   y=percentage, 
                   fill=meta25))+
  geom_bar(width = 1, stat = "identity")+
  scale_fill_manual(values = CATALYST:::.cluster_cols[1:25])+coord_polar("y",start=0)+
  blank_theme+theme(axis.text.x=element_blank())+
    guides(fill=guide_legend(title="FlowSOM metacluster",override.aes = list(size = 4))))

markers_to_plot = c("IFN_pos","TNF_pos","CD107a_pos")
df_plot = df_final[(df_final$marker %in% markers_to_plot &df_final$stimulation=="GAG"),c("meta25","marker","n_pos_backgroundsubtracted")]

# transform n_pos_backgroundsubtracted to a percentage 
df_plot = df_plot %>% group_by(marker) %>%
  mutate(percentage = n_pos_backgroundsubtracted / sum(n_pos_backgroundsubtracted) * 100)

fig1e_alt = print(ggplot(df_plot, 
                         aes(x="", 
                             y=percentage, 
                             fill=meta25))+
                    geom_bar(width = 1, stat = "identity")+
                    facet_wrap(~marker,
                               ncol = 1,
                               labeller = as_labeller(c("CD107a_pos" = "CD107a",
                                                        "IFN_pos" = "IFN",
                                                        "TNF_pos" = "TNF")))+
                    scale_fill_manual(values = CATALYST:::.cluster_cols[1:25])+coord_polar("y",start=0)+
                    blank_theme+theme(axis.text.x=element_blank(),legend.position = "none"))

# design <- "
# AAAABBBB
# AAAABBBB
# CCCCDDDD
# CCCCDDDD
# CCCCEEE#
# CCCCEEE#
# "
# png(filename = glue("{dir_out}/Figure1.png"), width = 16,height = 15,unit = "in",res = 300)
# fig1a + fig1b + fig1c + fig1d + fig1e + plot_layout(design = design)+ plot_annotation(tag_levels = 'A')&theme(plot.tag = element_text(size = 20, face = 'bold'))
# dev.off()


design <- "
AAAAABBBBB
AAAAABBBBB
AAAAABBBBB
CCCCCDDDEE
CCCCCDDDEE
CCCCCDDDEE
CCCCCDDDEE
"

fig1a$layers[[1]]$aes_params$size = 0.005
fig1b$layers[[1]]$aes_params$size = 0.005
fig1c$layers[[1]]$aes_params$size = 0.000000001
fig1d_alt$layers[[1]]$aes_params$size = 0.000000001


png(filename = glue("{dir_out}/Figure1_alt.png"), width = 16,height = 15,unit = "in",res = 300)
fig1a + fig1b + fig1c + fig1d_alt + fig1e_alt + plot_layout(design = design)+ plot_annotation(tag_levels = 'A')&theme(plot.tag = element_text(size = 20, face = 'bold'))
dev.off()

```


## 6) Session info

```{r Session info, message = TRUE}

sessionInfo()

```
